---
title: "Varying temporal trends in the levels of six groups of legacy persistent organic pollutants (POPs) in liver of three gadoid species from the North Sea: Data treatment and statistical analysis"
output:
  bookdown::html_document2:
    theme: journal
    code_folding: hide
    number_sections: false
editor_options:
  markdown:
    wrap: 150
always_allow_html: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  echo = TRUE,
  warning = FALSE
)
library(tidyverse)
library(readxl)
library(janitor)
library(broom)
library(scales)
library(visreg)
library(jtools)
library(ggpp)
library(modelr)
library(gt)
library(car)
library(ggh4x)
theme_set(theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.key = element_blank(),
    strip.background = element_rect(colour="black", fill="white"),
    strip.text.x = element_text(face = "bold"),
    strip.text.y = element_text(face = "bold")
  ))
```

## Data cleaning

```{r}
  cleaned_data_w_age <- read_excel("data/20230424_nordsjoen_stepan.xlsx", skip = 1) %>%
    clean_names() %>% 
    select(species, year, alder_y, length_cm, fettinnhold_percent, lsi_percent, hel_vekt_g, lever_vekt_g, matches("w_w")) %>%
    drop_na(species, year, length_cm, fettinnhold_percent, lsi_percent) %>%
    mutate(across(.cols = -species, .fn = ~ as.numeric(.x))) %>% 
    # log transform response variables and covariate length
    mutate(across(.cols = matches("w_w"), .fn = ~ log10(as.numeric(.x) + 1))) %>%
    mutate(across(.cols = matches("length_cm"), .fn = ~ log10(as.numeric(.x)))) %>%
    mutate(across(.cols = matches("fettinnhold_percent|year|lsi|vekt_g|alder"), .fn = ~ as.numeric(.x))) %>%
    rename_with(
      ~ str_replace_all(.x, c(
        "w_w" = "ww",
        "^s" = "",
        "fettinnhold_percent" = "fat_pc",
        "lsi_percent" = "lsi_pc"
      )) %>%
        str_to_lower(),
      .cols = !matches("species")
    ) %>%
    # remove obs with no POP data
    filter(if_any(matches("ww"), ~ !is.na(.x))) %>%
    pivot_longer(cols = matches("ww"), names_to = "variable", values_to = "value") %>%
    # update two values found to be incorrectly entered
    mutate(
      hel_vekt_g = if_else(hel_vekt_g == 635 & lever_vekt_g == 151.9, 6135, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 6135 & lever_vekt_g == 151.9, lever_vekt_g / hel_vekt_g, lsi_pc),
      hel_vekt_g = if_else(hel_vekt_g == 1004 & lever_vekt_g == 780.0, 10040, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 10040 & lever_vekt_g == 780.0, lever_vekt_g / hel_vekt_g, lsi_pc),
    )


  raw_data <- read_excel("data/20230424_nordsjoen_stepan.xlsx", skip = 1) %>%
    clean_names() %>%
    select(species, year, length_cm, fettinnhold_percent, lsi_percent, hel_vekt_g, lever_vekt_g, matches("w_w")) %>%
    drop_na(species, year, length_cm, fettinnhold_percent, lsi_percent) %>%
    mutate(across(.cols = -species, .fn = ~ as.numeric(.x)))

  cleaned_data <- raw_data %>%
    # log transform response variables and covariate length
    mutate(across(.cols = matches("w_w"), .fn = ~ log10(as.numeric(.x) + 1))) %>%
    mutate(across(.cols = matches("length_cm"), .fn = ~ log10(as.numeric(.x)))) %>%
    mutate(across(.cols = matches("fettinnhold_percent|year|lsi|vekt_g"), .fn = ~ as.numeric(.x))) %>%
    rename_with(
      ~ str_replace_all(.x, c(
        "w_w" = "ww",
        "^s" = "",
        "fettinnhold_percent" = "fat_pc",
        "lsi_percent" = "lsi_pc"
      )) %>%
        str_to_lower(),
      .cols = !matches("species")
    ) %>%
    # remove obs with no POP data
    filter(if_any(matches("ww"), ~ !is.na(.x))) %>%
    pivot_longer(cols = matches("ww"), names_to = "variable", values_to = "value") %>%
    # update two values found to be incorrectly entered
    mutate(
      hel_vekt_g = if_else(hel_vekt_g == 635 & lever_vekt_g == 151.9, 6135, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 6135 & lever_vekt_g == 151.9, lever_vekt_g / hel_vekt_g, lsi_pc),
      hel_vekt_g = if_else(hel_vekt_g == 1004 & lever_vekt_g == 780.0, 10040, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 10040 & lever_vekt_g == 780.0, lever_vekt_g / hel_vekt_g, lsi_pc),
    )

```


## Data description

```{r}
cleaned_data %>%
  group_by(species, variable, year) %>%
  summarise(n_obs = n()) %>%
  ungroup() %>%
  select(-variable) %>%
  distinct(species, year, .keep_all = TRUE) %>% pivot_wider(names_from = "species", values_from = "n_obs") %>% 
  gt(
    caption = "Samples per year per species.",
    groupname_col = "species",
    rowname_col = "year"
  )
```


* Observations:
  + Approx. 250 for each fish species (3) for each contaminant (6).
  + For each year at least 41 observations for each species and contaminant except for cod in 2019, with only 9 observations.
* Response variables:
  + Concentrations of six different POPs in liver in ug per kg liver wet weight.
* Explanatory variables:
  + species (cod, haddock or saithe)
  + year (in which fish was caught)
  + weight of fish (in g, omitted)
  + weight of liver (in g, omitted)
  + length of fish (in cm, log transformed for normality)
  + fat percentage (percentage of liver weight)
  + lsi (liver somatic index or hepatosomatic index: liver weight vs total weight, nutritional condition)
  + fish age (in years, omitted)

From previous studies it was known that effect of these variables will differ between species, owing to their distinct physiology and feeding habits. Hence, a model was fitted to each species as comparisons between different fish species were not a priority. Known confounders are fish age or proxies such as fish length and fish weight as well as liver weight, liver fat content and hepatosomatic or liver somatic index.

## Model selection

To investigate the hypothesis, "*is there a time trend for any of the contaminants in any of the species?*", one identical model was chosen and fitted to each different dataset (3 species x  6 contaminants = 18). This was justified as including an additional explanatory variable should be inconsequential due to the large number of observations. Due to collinearity with lsi and fish length, weight of liver and fish was omitted. Fish age was omitted due to missingness of a substantial fraction of observations and collinearity with fish length. As such, a linear model including explanatory variables year, fish length, fat percentage and lsi was chosen.


## Model fitting

```{r model}
# linear regression model
pop_lm <- function(in_dat) {
  lm(
    data = in_dat, formula = value ~ year + length_cm + fat_pc + lsi_pc,
    na.action = na.exclude
  )
}

```


```{r model-fit-partial}

# conditional partial residuals manually calculated
p_man <- cleaned_data %>%
  ungroup() %>%
  group_by(species, variable) %>%
  drop_na(value) %>%
  nest() %>%
  # apply model and calculate medians of covariates
  mutate(
    model = map(data, pop_lm),
    med_length_cm = map(
      data,
      ~ median(.x$length_cm, na.rm = TRUE)
    ) %>% as.numeric(),
    med_fat_pc = map(
      data,
      ~ median(.x$fat_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    med_lsi_pc = map(
      data,
      ~ median(.x$lsi_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_length_cm = map(
      data,
      ~ mean(.x$length_cm, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_fat_pc = map(
      data,
      ~ mean(.x$fat_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_lsi_pc = map(
      data,
      ~ mean(.x$lsi_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_year = map(
      data,
      ~ mean(.x$year, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_value = map(
      data,
      ~ mean(.x$value, na.rm = TRUE)
    ) %>% as.numeric()
  ) %>%
  mutate(
    glance = map(model, broom::glance),
    year_pval = map(model, ~ broom::tidy(.x)[[2, 5]] %>% as.numeric()),
    year_estimate = map(model, ~ broom::tidy(.x)[[2, 2]] %>% as.numeric() %>% {10^. - 1}),
    year_estimate_nolog = map(model, ~ broom::tidy(.x)[[2, 2]] %>% as.numeric()),
    # extracting model coefficients for partial residual calculation
    tidy_coeff = map(model, ~ broom::tidy(.x) %>%
      select(term, estimate) %>%
      pivot_wider(names_from = "term", values_from = "estimate") %>%
      clean_names() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_coeff"))),
    # tidy gets summary variables of the model coefficients
    tidy = map(model, ~ broom::tidy(.x)),
    # augment gets per observation variables (e.g. residuals)
    augment = map2(model, data, ~ broom::augment(.x, .y)),
    partial_resids = map(model, ~ residuals(.x, type = "partial") %>%
      as_tibble() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_partial_resid"))),
    normal_resids = map(model, ~ residuals(.x) %>%
      as_tibble() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_normal_resid")))
  ) %>%
  # add package-generated plots for an extra check
  mutate(
    visreg = map(
      model,
      ~ visreg(.x, "year", gg = TRUE, type = "conditional")
    ),
    effect_plot = map(
      model,
      ~ effect_plot(.x, pred = year, interval = TRUE, partial.residuals = TRUE)
    ),
    variable = toupper(variable) %>% str_replace_all("_WW", ""),
    
  ) %>%
  # vifs to check for collinearity
  mutate(
    vifs = map(model, ~ car::vif(.x) %>% as_tibble())
    )

```

## Model validation

### Linearity

```{r}

#Residuals vs fitted

p_man %>% unnest(augment) %>%
  ggplot(aes(.fitted, .resid)) +
    geom_point(size = 0.2) + 
    geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")

```

Some non-linear patterns observed e.g for HCB in haddock and cod, yet patterns are not systematic across species and contaminants. Thus, the current model was retained.

```{r}

#Residuals vs each covariate

p_man %>% unnest(augment) %>%
  ggplot(aes(length_cm, .resid)) +
    geom_point(size = 0.2) + 
      geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")

p_man %>% unnest(augment) %>%
  ggplot(aes(fat_pc, .resid)) +
    geom_point(size = 0.2) + 
      geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")

p_man %>% unnest(augment) %>%
  ggplot(aes(lsi_pc, .resid)) +
    geom_point(size = 0.2) + 
    geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")


```



### Normality

#### QQ-plot of response

```{r}
ggplot(p_man %>% unnest(augment), aes(sample = value)) +
  stat_qq(size = 0.2) +
  stat_qq_line(color = "red", size = 0.4) +
  facet_grid(species ~ variable, scales = "free")
```

#### Distribution of response

```{r}
ggplot(p_man %>% unnest(augment), aes(value)) +
    geom_histogram(aes(y=..density..), color='gray50',
        alpha=0.2, position = "identity")+
    geom_density(alpha=0.2, fill = "red")+
  facet_grid(species ~ variable, scales = "free")
```


#### Shapiro-Wilk's test


```{r shapiro-3col}
shap_table <- p_man %>% unnest(augment) %>% group_by(species, variable) %>% summarise(shapiro = shapiro.test(.std.resid) %>% tidy()) %>% unnest(shapiro) %>% select(-method, -statistic)

fish_table <- function(x){
  gt(x) %>% 
    tab_style(
    location = cells_body(
      columns = p.value,
      rows = p.value < 0.05
    ),
    style = list(cell_text(color = "red"))
  ) %>% fmt_number(columns = 2, decimals = 3) %>% 
    tab_options(column_labels.hidden = TRUE) %>% 
    as_raw_html() # return as html
}

left_table <- 
  shap_table %>% 
  filter(species == "Cod") %>% 
  ungroup() %>% 
  select(-species) %>% 
  fish_table()

mid_table <- 
  shap_table %>% 
  filter(species == "Haddock") %>%
    ungroup() %>% 
  select(-species) %>% 
  fish_table()

right_table <- 
  shap_table %>% 
  filter(species == "Saithe") %>%
    ungroup() %>% 
  select(-species) %>% 
  fish_table()

data_tables <- data.frame(left_table = left_table,
                          mid_table = mid_table,
                          right_table = right_table)

data_tables %>% 
  gt(
    caption = "Shapiro-Wilk's test."
  ) %>% 
  fmt_markdown(columns = TRUE) %>% #render cell contents as html
  cols_label(left_table = "Cod",
             mid_table = "Haddock",
             right_table = "Saithe") %>% 
  tab_footnote("Red text indicates p-value < 0.05.")

```

Deviation from normality is observed and p-values should be interpreted with caution. Bimodality observed for HCH could not be linked to sampling location.

### Homoskedasticity

#### Scale-location plot

```{r}
p_man %>%
  unnest(augment) %>%
  ggplot(aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point(size = 0.2) +
  geom_smooth(method = "loess", formula = y ~ x, span = 2, se = T, size = 0.5) +
  geom_smooth(method = "lm", formula = y ~ 1, se = FALSE, color = "red", size = 0.5, linetype="dashed", span = 2)+
  facet_grid(species ~ variable, scales = "free")
```

No clear systematic patterns in the variance versus fitted value is observed, most showing relatively homoskedastic behavior.

### Collinearity

```{r}
# year
cleaned_data %>%
  select(-variable, -value, -hel_vekt_g, -lever_vekt_g) %>%
  mutate(across(!matches("species"), ~ as.numeric(.x))) %>%
  pivot_longer(cols = -c("species", "year"), names_to = "variable", values_to = "value") %>%
  ggplot(aes(as_factor(year), value)) +
  geom_point() + geom_boxplot()+
  theme(
      axis.text.x = element_text(angle = 90, hjust = 1)
    )+
  ggh4x::facet_grid2(species ~ variable, scales = "free_y", independent = "y")

# # fat_pc
# cleaned_data %>%
#   select(-variable, -value, -hel_vekt_g, -lever_vekt_g) %>%
#   mutate(across(!matches("species"), ~ as.numeric(.x))) %>%
#   pivot_longer(cols = -c("species", "year"), names_to = "variable", values_to = "value") %>%
#   ggplot(aes(as_factor(year), value)) +
#   geom_point() + geom_boxplot()+
#   theme(
#       axis.text.x = element_text(angle = 90, hjust = 1)
#     )+
#   ggh4x::facet_grid2(species ~ variable, scales = "free_y", independent = "y")
# 
# # lsi_pc
# cleaned_data %>%
#   select(-variable, -value, -hel_vekt_g, -lever_vekt_g) %>%
#   mutate(across(!matches("species"), ~ as.numeric(.x))) %>%
#   pivot_longer(cols = -c("species", "year"), names_to = "variable", values_to = "value") %>%
#   ggplot(aes(as_factor(year), value)) +
#   geom_point() + geom_boxplot()+
#   theme(
#       axis.text.x = element_text(angle = 90, hjust = 1)
#     )+
#   ggh4x::facet_grid2(species ~ variable, scales = "free_y", independent = "y")
# 
# # length_cm
# cleaned_data %>%
#   select(-variable, -value, -hel_vekt_g, -lever_vekt_g) %>%
#   mutate(across(!matches("species"), ~ as.numeric(.x))) %>%
#   pivot_longer(cols = -c("species", "year"), names_to = "variable", values_to = "value") %>%
#   ggplot(aes(as_factor(year), value)) +
#   geom_point() + geom_boxplot()+
#   theme(
#       axis.text.x = element_text(angle = 90, hjust = 1)
#     )+
#   ggh4x::facet_grid2(species ~ variable, scales = "free_y", independent = "y")
```

VIFs were determined to be below 3 for each dependent variable. However, some collinearity e.g. a decrease in LSI for haddock with year due to fish weight/size increase is observed.



### Outliers

```{r}
# cleaned_data %>%
#   select(-variable, -value) %>%
#   mutate(across(!matches("species"), ~ as.numeric(.x))) %>%
#   pivot_longer(cols = -"species", names_to = "variable", values_to = "value") %>%
#   ggplot(aes(value)) +
#   geom_histogram(bins = 50) +
#   facet_grid(species ~ variable, scales = "free")+ theme(
#       axis.text.x = element_text(angle = 90, hjust = 1)
#     )
# 
# cleaned_data %>%
#   select(-variable, -value) %>%
#   mutate(across(!matches("species"), ~ as.numeric(.x))) %>%
#   pivot_longer(cols = -"species", names_to = "variable", values_to = "value") %>%
#   ggplot(aes(value)) + geom_boxplot() + facet_grid(species ~ variable, scales = "free")+ theme(
#       axis.text.x = element_text(angle = 90, hjust = 1)
#     )
```
Two outliers were identified in LSI. Upon further investigation, it was determined that these values were the result of typographic errors during data collection. These errors were corrected during the data cleaning above.

## PCB and HCB levels

```{r pcb-hcb}
  pcb_hcb <- read_excel("data/20230907_Datasettet med PCB6.xlsx", skip = 1) %>%
    clean_names() %>% 
    select(species, year, alder_y, length_cm, fettinnhold_percent, lsi_percent, hel_vekt_g, lever_vekt_g, matches("w_w|pcb")) %>%
    drop_na(species, year, length_cm, fettinnhold_percent, lsi_percent) %>%
    mutate(across(.cols = -species, .fn = ~ as.numeric(.x))) %>% 
    mutate(across(.cols = matches("length_cm"), .fn = ~ as.numeric(.x))) %>%
    mutate(across(.cols = matches("fettinnhold_percent|year|lsi|vekt_g|alder|pcb"), .fn = ~ as.numeric(.x))) %>%
    rename_with(
      ~ str_replace_all(.x, c(
        "w_w" = "ww",
        "^s" = "",
        "fettinnhold_percent" = "fat_pc",
        "lsi_percent" = "lsi_pc"
      )) %>%
        str_to_lower(),
      .cols = !matches("species")
    ) %>%
    # remove obs with no POP data
    filter(if_any(matches("ww|pcb"), ~ !is.na(.x))) %>%
    pivot_longer(cols = matches("ww|pcb"), names_to = "variable", values_to = "value") %>%
    # update two values found to be incorrectly entered
    mutate(
      hel_vekt_g = if_else(hel_vekt_g == 635 & lever_vekt_g == 151.9, 6135, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 6135 & lever_vekt_g == 151.9, lever_vekt_g / hel_vekt_g, lsi_pc),
      hel_vekt_g = if_else(hel_vekt_g == 1004 & lever_vekt_g == 780.0, 10040, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 10040 & lever_vekt_g == 780.0, lever_vekt_g / hel_vekt_g, lsi_pc),
    )




pcb_hcb %>%
  filter(variable %in% c("pcb6", "hcb_ww")) %>% ungroup() %>% 
  mutate(hline_leg = if_else(variable == "pcb6", 200, 10) %>%
           as.numeric()) %>% 
  # unnest(data) %>%
  # mutate(value = 10^value - 1) %>%
  filter(value < 1500) %>%
  ggplot(aes(year, value, group = year)) +
  geom_jitter(width = 0.35, size = 0.3) +
  geom_boxplot(outlier.shape = NA, width = 1.65) +
  scale_x_continuous(breaks = c(2019, 2016, 2013, 2011, 2008, 2005, 2010)) +
  stat_summary(fun.y = mean, geom = "point", shape = 3, size = 1.8, color = "red", fill = "red") +
  geom_hline(na.rm = TRUE, aes(yintercept = hline_leg), color = "red", linetype = "dashed") +
    scale_y_continuous(breaks = pretty_breaks(n = 6))+

  ylab("Concentration [\u00b5g/kg]") +
  xlab("Year") +
  facet_grid(variable ~ species, scales = "free")

ggsave("hcb_pcb_reg.svg", width = 6, height = 3.5)
```

## Bivariate correlations

### LSI DDTs and HCHs in Haddock 

```{r bivar-cor}
p_man %>% 
  filter(variable %in% c("DDT", "HCH"),
         species == "Haddock") %>% unnest(data) %>% 
  mutate(value = 10^value - 1) %>%
  ggplot(aes(lsi_pc, value))+
  geom_point(size = 0.5) + 
  geom_smooth(method = "lm", color = "red", size = 0.75) + 
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))+
  facet_wrap(~variable, scales = "free") + 
    ylab("Concentration [\u00b5g/kg]") +
  xlab("LSI [%]")
ggsave("ddt_hch_bivar.svg", width = 9, height = 3)
```

### PCB7 and DDT versus age

```{r}
cleaned_data_w_age %>%
  mutate(
    variable = toupper(variable) %>% str_replace_all("_WW", "")
  ) %>% 
  filter(variable %in% c("DDT", "PCB7"),
         species == "Cod") %>% 
  mutate(value = 10^value - 1) %>%
  ggplot(aes(alder_y, value))+
  geom_point(size = 0.5) + 
  geom_smooth(method = "lm", color = "red", size = 0.75) + 
  scale_x_continuous(breaks = pretty_breaks(n = 5))+
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))+
  facet_wrap(~factor(variable, levels = c(
    "PCB7",
    "DDT",
    "HCH",
    "HCB",
    "TNC",
    "PBDE"
  )), scales = "free",
             ) + 
    ylab("Concentration [\u00b5g/kg]") +
  xlab("Age [years]")
ggsave("pcb_ddt_age.svg", width = 9, height = 3)
  
```


## Partial residuals

### Background

Assuming the least square estimate of the covariates $X_{1}$, $X_{2}$ and $X_{3}$ and $y_i$ and $\epsilon_i$ is the response and residual for the i-th point:

$$
y_i = f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}) + \epsilon_i.
$$

The partial residual of $X_3$ for the i-th observation is then generally defined as (Sohil et al., 2021) (Larsen and McCleary, 1972):

$$
r_i = y_i-f_1\left(x_{i 1}\right)-f_2\left(x_{i 2}\right) = \epsilon_i + f_3(x_{i3}) ,
$$
 
 where in the present case


$$
y_i = \hat b_0 + \hat b_{year}(year_{i})+ \hat b_{length}(length_{i})+\hat b_{fat pc}(fatpc_{i})+ \hat b_{lsi pc}(lsi pc_{i})+ \epsilon_i
$$
or, similarly

$$
 y_i = \bar y + \hat b_{year}\cdot (x_{year}-\bar x_{year}) + \hat b_{length}\cdot (x_{length}-\bar x_{length}) + \hat b_{fatpc}\cdot (x_{fatpc}-\bar x_{fatpc})  + \hat b_{lsipc}\cdot (x_{lsipc}-\bar x_{lsipc}) + \epsilon_i
$$

### Conditional partial residual plots

```{r}
p_man %>%
  unnest(tidy_coeff, partial_resids, data, normal_resids, augment) %>%
  ungroup() %>%
  ggplot(aes(year, (value_normal_resid + year * year_coeff + intercept_coeff + med_length_cm * length_cm_coeff + med_fat_pc * fat_pc_coeff + med_lsi_pc * lsi_pc_coeff))) +
  geom_jitter(size = 0.3, na.rm = T, height = 0, width = 0.25) +
  # geom_point(aes(year, value), size = 0.2, color = "darkgrey", position = position_jitternudge(width = 0.2, height = 0,
  #                                    seed = 123, x = 0.5,
  #                                    direction = "as.is",
  #                                    nudge.from = "jittered"))+
  ### TEST COND PART RESID
  # geom_point(aes(year, value_normal_resid + mean_value +
  #                year_coeff*(year - mean_year) + length_cm_coeff*(med_length_cm - mean_length_cm) + fat_pc_coeff*(med_fat_pc - mean_fat_pc)), size = 0.2, color = "orange", position = position_jitternudge(width = 0.2, height = 0,
  #                                    seed = 123, x = -0.5,
  #                                    direction = "as.is",
  #                                    nudge.from = "jittered"))+
  ### TEST COND PART RESID END
  stat_smooth(method = "lm", na.rm = TRUE, se = FALSE, color = "red", size = 0.5) +
  # conditional coloring percentage change
  geom_text(
   data = p_man %>% group_by(species, variable) %>%
     distinct(year_estimate, .keep_all = TRUE),
   color =
     case_when(
       p_man %>% group_by(species, variable) %>%
         distinct(year_estimate) %>% pull(year_estimate) > 0 &
         p_man %>%
           group_by(species, variable) %>%
           distinct(year_pval) %>%
           pull(year_pval) < 0.05 ~ "red",
       p_man %>% group_by(species, variable) %>%
         distinct(year_estimate) %>% pull(year_estimate) < 0 &
         p_man %>%
           group_by(species, variable) %>%
           distinct(year_pval) %>%
           pull(year_pval) < 0.05 ~ "green",
       TRUE ~ "black"
     ),
   fontface = "bold",
   x = -Inf, y = Inf, hjust = -0.075, vjust = 1.5,
   size = p_man %>% group_by(species, variable) %>%
         distinct(year_estimate) %>% pull(year_estimate) %>% as.numeric() %>% abs()*150/14 + 2.3,
   # size = 3.7,
   aes(label = paste0(round(year_estimate %>% as.numeric() * 100, digits = 1) %>% number(style_positive = "plus", accuracy = 0.1), "%"))
 )+
  # p-values
    geom_text(
    data = p_man %>% group_by(species, variable) %>%
      distinct(year_pval, .keep_all = TRUE), color = "grey30",
    x = 2004.9, y = 0.1, hjust = 0, vjust = 0.6, size = 2.5,
    aes(label = paste0("p = ", round(year_pval %>% as.numeric(), digits = 3) %>% format(nsmall = 2)))
  ) +
  
  #   geom_text(
  #   data = p_man %>% unnest(tidy) %>% group_by(species, variable) %>%
  #     distinct(year_coeff, .keep_all = TRUE), color = "red",
  #   x = -Inf, y = Inf, hjust = -0.1, vjust = 3, size = 2.5,
  #   aes(label = paste0("slope = ", round(year_coeff %>% as.numeric(), digits = 4)))
  # ) +
  scale_x_continuous(breaks = c(2019, 2016, 2013, 2011, 2008, 2005, 2010)) +
  scale_y_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3), limits = c(0, 3.2)) +
  ylab("log(Concentration + 1) [\u00b5g/kg]") +
  xlab("Year | (Length, Fat and LSI)") +

    facet_grid(species ~ factor(variable, levels = c(
    "PCB7",
    "DDT",
    "HCH",
    "HCB",
    "TNC",
    "PBDE"
  )))

ggsave("conditional_pr_stepan_data.svg", width = 11, height = 6)

# comditional jtools and visreg
# p_man$visreg[[18]] + scale_y_continuous(breaks = pretty_breaks(n = 10))
# p_man$effect_plot[[18]]
# patchwork::wrap_plots(p_man$visreg)
```

## Strength of decrease

### Table

```{r}

p_man %>% clean_names() %>%
  select(species, variable, tidy) %>%
  unnest(tidy) %>%
  filter(term == "year") %>%
  select(species, variable, p.value, estimate) %>%
  mutate(estimate = 10^estimate - 1) %>%
  gt(caption = "Percentage decrease per year.",
     groupname_col = "variable",
     rowname_col = "species") %>% 
  tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate > 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "red"))
  ) %>% 
  
    tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate < 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "green"))
  ) %>% 
  fmt_number(everything(), decimals = 3) %>% 
  fmt_percent(estimate, decimals = 1) %>% 
  tab_footnote("Estimated annual percentage change obtained by exponentiation and corresponding p-values. Green color represents a significant decrease at the p = 0.05 level, while red color indicates a significant increase.")

```

```{r lsi-fat-str-of-change}
p_man %>% clean_names() %>%
  select(species, variable, tidy) %>%
  unnest(tidy) %>% 
  filter(term == "lsi_pc") %>%
  select(species, variable, p.value, estimate) %>%
  mutate(estimate = 10^estimate - 1) %>%
  gt(caption = "Percentage change per percentage increase in lsi",
     groupname_col = "variable",
     rowname_col = "species") %>% 
  tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate > 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "red"))
  ) %>% 
  
    tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate < 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "green"))
  ) %>% 
  fmt_number(everything(), decimals = 3) %>% 
  fmt_percent(estimate, decimals = 1) %>% 
  tab_footnote("Estimated percentage change per percent LSI increase obtained by exponentiation and corresponding p-values. Green color represents a significant decrease at the p = 0.05 level, while red color indicates a significant increase.")

p_man %>% clean_names() %>%
  select(species, variable, tidy) %>%
  unnest(tidy) %>% 
  filter(term == "fat_pc") %>%
  select(species, variable, p.value, estimate) %>%
  mutate(estimate = 10^estimate - 1) %>%
  gt(caption = "Percentage change per percentage increase in fat",
     groupname_col = "variable",
     rowname_col = "species") %>% 
  tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate > 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "red"))
  ) %>% 
  
    tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate < 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "green"))
  ) %>% 
  fmt_number(everything(), decimals = 3) %>% 
  fmt_percent(estimate, decimals = 1) %>% 
  tab_footnote("Estimated percentage change per percent fat increase obtained by exponentiation and corresponding p-values. Green color represents a significant decrease at the p = 0.05 level, while red color indicates a significant increase.")

```


### Visualization without log-transformation

```{r}

cleaned_data_notrans <- raw_data %>%
    # log transform covariate length
    mutate(across(.cols = matches("w_w"), .fn = ~ as.numeric(.x))) %>%
    mutate(across(.cols = matches("length_cm"), .fn = ~ log10(as.numeric(.x)))) %>%
    mutate(across(.cols = matches("fettinnhold_percent|year|lsi|vekt_g"), .fn = ~ as.numeric(.x))) %>%
    rename_with(
      ~ str_replace_all(.x, c(
        "w_w" = "ww",
        "^s" = "",
        "fettinnhold_percent" = "fat_pc",
        "lsi_percent" = "lsi_pc"
      )) %>%
        str_to_lower(),
      .cols = !matches("species")
    ) %>%
    # remove obs with no POP data
    filter(if_any(matches("ww"), ~ !is.na(.x))) %>%
    pivot_longer(cols = matches("ww"), names_to = "variable", values_to = "value") %>%
    # update two values found to be incorrectly entered
    mutate(
      hel_vekt_g = if_else(hel_vekt_g == 635 & lever_vekt_g == 151.9, 6135, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 6135 & lever_vekt_g == 151.9, lever_vekt_g / hel_vekt_g, lsi_pc),
      hel_vekt_g = if_else(hel_vekt_g == 1004 & lever_vekt_g == 780.0, 10040, hel_vekt_g),
      lsi_pc = if_else(hel_vekt_g == 10040 & lever_vekt_g == 780.0, lever_vekt_g / hel_vekt_g, lsi_pc),
    )

p_man_notrans <- cleaned_data_notrans %>%
  ungroup() %>%
  group_by(species, variable) %>%
  drop_na(value) %>%
  nest() %>%
  # apply model and calculate medians of covariates
  mutate(
    model = map(data, pop_lm),
    med_length_cm = map(
      data,
      ~ median(.x$length_cm, na.rm = TRUE)
    ) %>% as.numeric(),
    med_fat_pc = map(
      data,
      ~ median(.x$fat_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    med_lsi_pc = map(
      data,
      ~ median(.x$lsi_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_length_cm = map(
      data,
      ~ mean(.x$length_cm, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_fat_pc = map(
      data,
      ~ mean(.x$fat_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_lsi_pc = map(
      data,
      ~ mean(.x$lsi_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_year = map(
      data,
      ~ mean(.x$year, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_value = map(
      data,
      ~ mean(.x$value, na.rm = TRUE)
    ) %>% as.numeric()
  ) %>%
  mutate(
    glance = map(model, broom::glance),
    year_pval = map(model, ~ broom::tidy(.x)[[2, 5]] %>% as.numeric()),
    year_estimate = map(model, ~ broom::tidy(.x)[[2, 2]] %>% as.numeric() %>% {10^. - 1}),
    year_estimate_nolog = map(model, ~ broom::tidy(.x)[[2, 2]] %>% as.numeric()),
    # extracting model coefficients for partial residual calculation
    tidy_coeff = map(model, ~ broom::tidy(.x) %>%
      select(term, estimate) %>%
      pivot_wider(names_from = "term", values_from = "estimate") %>%
      clean_names() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_coeff"))),
    # tidy gets summary variables of the model coefficients
    tidy = map(model, ~ broom::tidy(.x)),
    # augment gets per observation variables (e.g. residuals)
    augment = map2(model, data, ~ broom::augment(.x, .y)),
    partial_resids = map(model, ~ residuals(.x, type = "partial") %>%
      as_tibble() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_partial_resid"))),
    normal_resids = map(model, ~ residuals(.x) %>%
      as_tibble() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_normal_resid")))
  ) %>%
  # add package-generated plots for an extra check
  mutate(
    visreg = map(
      model,
      ~ visreg(.x, "year", gg = TRUE, type = "conditional")
    ),
    effect_plot = map(
      model,
      ~ effect_plot(.x, pred = year, interval = TRUE, partial.residuals = TRUE)
    ),
    variable = toupper(variable) %>% str_replace_all("_WW", ""),
    
  ) %>%
  # vifs to check for collinearity
  mutate(
    vifs = map(model, ~ car::vif(.x) %>% as_tibble())
    )

p_man_notrans %>%
  unnest(tidy_coeff, partial_resids, data, normal_resids, augment) %>%
  ungroup() %>%
  ggplot(aes(year, (value_normal_resid + year * year_coeff + intercept_coeff + med_length_cm * length_cm_coeff + med_fat_pc * fat_pc_coeff + med_lsi_pc * lsi_pc_coeff))) +
  geom_jitter(size = 0.3, na.rm = T, height = 0, width = 0.25) +
  # geom_point(aes(year, value), size = 0.2, color = "darkgrey", position = position_jitternudge(width = 0.2, height = 0,
  #                                    seed = 123, x = 0.5,
  #                                    direction = "as.is",
  #                                    nudge.from = "jittered"))+
  ### TEST COND PART RESID
  # geom_point(aes(year, value_normal_resid + mean_value +
  #                year_coeff*(year - mean_year) + length_cm_coeff*(med_length_cm - mean_length_cm) + fat_pc_coeff*(med_fat_pc - mean_fat_pc)), size = 0.2, color = "orange", position = position_jitternudge(width = 0.2, height = 0,
  #                                    seed = 123, x = -0.5,
  #                                    direction = "as.is",
  #                                    nudge.from = "jittered"))+
  ### TEST COND PART RESID END
  geom_smooth(method = "lm", color = "orange")+
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.key = element_blank(),
    strip.background = element_rect(colour="black", fill="white"),
    strip.text.x = element_text(face = "bold"),
    strip.text.y = element_text(face = "bold")
  ) +
  scale_x_continuous(breaks = c(2019, 2016, 2013, 2011, 2008, 2005, 2010)) +
  ylab("Concentration [\u00b5g/kg]") +
  xlab("Year | (Length, Fat and LSI)") +

    facet_grid2(species ~ factor(variable, levels = c(
    "PCB7",
    "DDT",
    "HCH",
    "HCB",
    "TNC",
    "PBDE"
  )), scales = "free_y", independent = "y")
```

## Regression tables


```{r using-gt-only-cod}


# Cod
Cod_reg_tab <- rbind(
  p_man %>%
    filter(species == "Cod") %>%
    ungroup() %>%
    select(variable, model, tidy, glance, augment) %>%
    unnest(tidy) %>%
    select(-model, -glance, -augment) %>%
    select(-std.error, -statistic) %>%
    pivot_longer(cols = c(estimate, p.value), names_to = "property") %>%
    pivot_wider(names_from = variable, values_from = value) %>%
    mutate(
      term = str_replace_all(term, c(
        "year" = "Year",
        "length_cm" = "Length",
        "fat_pc" = "Fat %",
        "lsi_pc" = "LSI %"
      ))
    ) %>%
    filter(!str_detect(term, "tercept")) %>%
    ungroup(),
  p_man %>%
    filter(species == "Cod") %>%
    ungroup() %>%
    select(variable, model, tidy, glance, augment) %>%
    unnest(glance) %>%
    select(-model, -tidy, -augment) %>%
    mutate(term = "Model") %>%
    select(variable, term, adj.r.squared, statistic, p.value, nobs) %>%
    rename(n_obs = nobs, F.statistic = statistic) %>%
    pivot_longer(cols = -c(variable, term), names_to = "property") %>%
    pivot_wider(names_from = "variable", values_from = "value")
) %>%
  gt(
    caption = "Regression summary table for Cod, bold indicating non-significant coefficients.",
    groupname_col = "term",
    rowname_col = "property"
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    decimals = 3
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    rows = property == "n_obs",
    decimals = 0
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    rows = property == "F.statistic",
    decimals = 0
  ) %>%
  tab_style(
    location = cells_body(
      columns = PCB7,
      rows = property == "p.value" & (PCB7 > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = DDT,
      rows = property == "p.value" & (DDT > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = HCH,
      rows = property == "p.value" & (HCH > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = HCB,
      rows = property == "p.value" & (HCB > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = TNC,
      rows = property == "p.value" & (TNC > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = PBDE,
      rows = property == "p.value" & (PBDE > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(
      columns = everything()
    )
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups(groups = everything())
  ) %>% tab_header("Cod")
  Cod_reg_tab
  gtsave(Cod_reg_tab, "Cod_reg_tab.docx")
  

```


```{r gt-only-haddock}
# Haddock
Haddock_reg_tab <- rbind(
  p_man %>%
    filter(species == "Haddock") %>%
    ungroup() %>%
    select(variable, model, tidy, glance, augment) %>%
    unnest(tidy) %>%
    select(-model, -glance, -augment) %>%
    select(-std.error, -statistic) %>%
    pivot_longer(cols = c(estimate, p.value), names_to = "property") %>%
    pivot_wider(names_from = variable, values_from = value) %>%
    mutate(
      term = str_replace_all(term, c(
        "year" = "Year",
        "length_cm" = "Length",
        "fat_pc" = "Fat %",
        "lsi_pc" = "LSI %"
      ))
    ) %>%
    filter(!str_detect(term, "tercept")) %>%
    ungroup(),
  p_man %>%
    filter(species == "Haddock") %>%
    ungroup() %>%
    select(variable, model, tidy, glance, augment) %>%
    unnest(glance) %>%
    select(-model, -tidy, -augment) %>%
    mutate(term = "Model") %>%
    select(variable, term, adj.r.squared, statistic, p.value, nobs) %>%
    rename(n_obs = nobs, F.statistic = statistic) %>%
    pivot_longer(cols = -c(variable, term), names_to = "property") %>%
    pivot_wider(names_from = "variable", values_from = "value")
) %>%
  gt(
    caption = "Regression summary table for Haddock, bold indicating non-significant coefficients.",
    groupname_col = "term",
    rowname_col = "property"
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    decimals = 3
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    rows = property == "n_obs",
    decimals = 0
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    rows = property == "F.statistic",
    decimals = 0
  ) %>%
  tab_style(
    location = cells_body(
      columns = PCB7,
      rows = property == "p.value" & (PCB7 > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = DDT,
      rows = property == "p.value" & (DDT > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = HCH,
      rows = property == "p.value" & (HCH > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = HCB,
      rows = property == "p.value" & (HCB > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = TNC,
      rows = property == "p.value" & (TNC > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = PBDE,
      rows = property == "p.value" & (PBDE > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(
      columns = everything()
    )
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups(groups = everything())
  ) %>% tab_header("Haddock")
  Haddock_reg_tab
  gtsave(Haddock_reg_tab, "Haddock_reg_tab.docx")
  
 
```

```{r gt-only-saithe}
 # Saithe
Saithe_reg_tab <- rbind(
  p_man %>%
    filter(species == "Saithe") %>%
    ungroup() %>%
    select(variable, model, tidy, glance, augment) %>%
    unnest(tidy) %>%
    select(-model, -glance, -augment) %>%
    select(-std.error, -statistic) %>%
    pivot_longer(cols = c(estimate, p.value), names_to = "property") %>%
    pivot_wider(names_from = variable, values_from = value) %>%
    mutate(
      term = str_replace_all(term, c(
        "year" = "Year",
        "length_cm" = "Length",
        "fat_pc" = "Fat %",
        "lsi_pc" = "LSI %"
      ))
    ) %>%
    filter(!str_detect(term, "tercept")) %>%
    ungroup(),
  p_man %>%
    filter(species == "Saithe") %>%
    ungroup() %>%
    select(variable, model, tidy, glance, augment) %>%
    unnest(glance) %>%
    select(-model, -tidy, -augment) %>%
    mutate(term = "Model") %>%
    select(variable, term, adj.r.squared, statistic, p.value, nobs) %>%
    rename(n_obs = nobs, F.statistic = statistic) %>%
    pivot_longer(cols = -c(variable, term), names_to = "property") %>%
    pivot_wider(names_from = "variable", values_from = "value")
) %>%
  gt(
    caption = "Regression summary table for Saithe, bold indicating non-significant coefficients.",
    groupname_col = "term",
    rowname_col = "property"
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    decimals = 3
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    rows = property == "n_obs",
    decimals = 0
  ) %>%
  fmt_number(
    columns = matches("[A-Z]{2}"),
    rows = property == "F.statistic",
    decimals = 0
  ) %>%
  tab_style(
    location = cells_body(
      columns = PCB7,
      rows = property == "p.value" & (PCB7 > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = DDT,
      rows = property == "p.value" & (DDT > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = HCH,
      rows = property == "p.value" & (HCH > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = HCB,
      rows = property == "p.value" & (HCB > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = TNC,
      rows = property == "p.value" & (TNC > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    location = cells_body(
      columns = PBDE,
      rows = property == "p.value" & (PBDE > 0.05)
    ),
    style = list(cell_text(weight = "bold"))
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(
      columns = everything()
    )
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups(groups = everything())
  ) %>% tab_header("Saithe")
  Saithe_reg_tab
  gtsave(Saithe_reg_tab, "Saithe_reg_tab.docx")
```

## Graphical abstract

```{r}
# p_man %>%
#   unnest(tidy_coeff, partial_resids, data, normal_resids, augment) %>%
#   ungroup() %>%
#   ggplot(aes(year, (value_normal_resid + year * year_coeff + intercept_coeff + med_length_cm * length_cm_coeff + med_fat_pc * fat_pc_coeff + med_lsi_pc * lsi_pc_coeff))) +
#   geom_jitter(size = 0.3, na.rm = T, height = 0, width = 0.25) +
#   stat_smooth(method = "lm", na.rm = TRUE, se = FALSE, color = "red", size = 0.5) +
#   # conditional coloring percentage change
#   geom_text(
#    data = p_man %>% group_by(species, variable) %>%
#      distinct(year_estimate, .keep_all = TRUE),
#    color =
#      case_when(
#        p_man %>% group_by(species, variable) %>%
#          distinct(year_estimate) %>% pull(year_estimate) > 0 &
#          p_man %>%
#            group_by(species, variable) %>%
#            distinct(year_pval) %>%
#            pull(year_pval) < 0.05 ~ "red",
#        p_man %>% group_by(species, variable) %>%
#          distinct(year_estimate) %>% pull(year_estimate) < 0 &
#          p_man %>%
#            group_by(species, variable) %>%
#            distinct(year_pval) %>%
#            pull(year_pval) < 0.05 ~ "green",
#        TRUE ~ "black"
#      ),
#    fontface = "bold",
#    x = -Inf, y = Inf, hjust = -0.075, vjust = 1.5,
#    size = p_man %>% group_by(species, variable) %>%
#          distinct(year_estimate) %>% pull(year_estimate) %>% as.numeric() %>% abs()*150/14 + 2.3,
#    # size = 3.7,
#    aes(label = paste0(round(year_estimate %>% as.numeric() * 100, digits = 1) %>% number(style_positive = "plus", accuracy = 0.1), "%"))
#  )+
#   # p-values
#     geom_text(
#     data = p_man %>% group_by(species, variable) %>%
#       distinct(year_pval, .keep_all = TRUE), color = "grey30",
#     x = 2004.9, y = 0.1, hjust = 0, vjust = 0.6, size = 2.5,
#     aes(label = paste0("p = ", round(year_pval %>% as.numeric(), digits = 3) %>% format(nsmall = 2)))
#   ) +
# 
#   scale_x_continuous(breaks = c(2019, 2016, 2013, 2011, 2008, 2005, 2010)) +
#   scale_y_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3), limits = c(0, 3.2)) +
#   ylab("log(Concentration + 1) [\u00b5g/kg]") +
#   xlab("Year | (Length, Fat and LSI)") +
#   
#   theme_bw() +
#   theme(
#     axis.text.x = element_text(angle = 90, hjust = 1),
#     legend.key = element_blank(),
#     strip.background = element_rect(colour="black", fill="white"),
#     strip.text.x = element_text(face = "bold"),
#     strip.text.y = element_text(face = "bold"))+
#   
#     facet_grid(species ~ factor(variable, levels = c(
#     "PCB7",
#     "DDT",
#     "HCH",
#     "HCB",
#     "TNC",
#     "PBDE"
#   )))
# 
# ggsave("graphical_abstr.svg", width = 11, height = 6)
```


## Stomach content of haddock and cod

```{r}
library(ggrepel)
library(see)
piedata <- read_excel("data/stomach_all_haddock_and_cod_all_stations_2011.xlsx", skip = 2) %>% 
  select(1, Haddock, Cod) %>%
  rename(c("Group" = 1)) %>%
  pivot_longer(cols = -Group, names_to = "Species") %>%
  group_by(Species) %>%
  mutate(value = value / sum(value)) %>% 
  mutate(
  csum = rev(cumsum(rev(value))),
  pos = value / 2 + lead(csum, 1),
  pos = if_else(is.na(pos), value / 2, pos),
  group_id = 15-row_number()
)

piedata %>%
  mutate(Group = fct_reorder(Group, group_id)) %>% 
  ggplot(aes(
    x = "",
    y = value,
    fill = Group
  )) +
  geom_col() +
  coord_polar(theta = "y") +
  # guides(fill = guide_legend(title = "Group")) +
  # theme_void() +
  facet_wrap(~Species)
ggsave("stommach_pie.svg", width = 9, height = 6)

# https://r-charts.com/part-whole/donut-chart-ggplot2/
```


## Method description

Statistical analyses for the determination of temporal trends were carried out in R version 4.2.1. Data with missing covariates or response variables were removed, and response variables were added one prior to logtransformation with base 10 to meet assumptions of homoscedasticity and normality while preserving zero-valued observations. The covariate length was also logtransformed. To correct for known confounders, a multiple linear regression model with the covariates year, length, fat percentage and liver somatic index was fitted to the cleaned and transformed data for each POP and species. To visually demonstrate the effect of year on the POP concentrations correcting for other dependent variables, the conditional partial residuals versus year was plotted, keeping length, fat and lsi constant at their median values. The significance of the regression coefficient for the year was determined using a Wald test to test the null hypothesis of no time trend. As detailed in the supplementary material, model validation revealed some departure from model assumptions. As such, minor and marginally significant effects were interpreted with caution.


## References

* Sohil, F., Sohali, M.U., Shabbir, J., 2021. An introduction to statistical learning with applications in R: by Gareth James, Daniela Witten, Trevor Hastie, and Robert Tibshirani, New York, Springer Science and Business Media, 2013, $41.98, eISBN: 978-1-4614-7137-7. Statistical Theory and Related Fields 1–1. https://doi.org/10.1080/24754269.2021.1980261

* Larsen, W.A., McCleary, S.J., 1972. The Use of Partial Residual Plots in Regression Analysis. Technometrics 14, 781–790. https://doi.org/10/gp2cxn
