---
title: "Levels and temporal trends of persistent organic pollutants (POPs) in liver of atlantic cod, haddock and saithe from the northeastern North Sea "
output:
  bookdown::html_document2:
    theme: journal
    code_folding: hide
    number_sections: false
editor_options:
  markdown:
    wrap: 150
always_allow_html: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  echo = TRUE,
  warning = FALSE
)
library(tidyverse)
library(readxl)
library(janitor)
library(broom)
library(scales)
library(visreg)
library(jtools)
library(ggpp)
library(gtsummary)
library(skimr)
library(leaps)
library(FactoMineR)
library(vcd)
library(factoextra)
library(modelr)
library(gt)
library(DataExplorer)
```


## Data exploration and cleaning


```{r}
  raw_data <- read_excel("data/20230424_nordsjoen_stepan.xlsx", skip = 1) %>%
    clean_names() %>% 
      select(species, year, length_cm, fettinnhold_percent, lsi_percent, hel_vekt_g, lever_vekt_g, matches("w_w")) %>%
    drop_na(species, year, length_cm, fettinnhold_percent, lsi_percent) %>% 
  mutate(across(.cols = -species, .fn = ~ as.numeric(.x)))

cleaned_data <- raw_data %>% 
  # log transform response variables and covariate length
    mutate(across(.cols = matches("w_w"), .fn = ~ log10(as.numeric(.x) + 1))) %>%
    mutate(across(.cols = matches("length_cm"), .fn = ~ log10(as.numeric(.x)))) %>%
    mutate(across(.cols = matches("fettinnhold_percent|year|lsi|vekt_g"), .fn = ~ as.numeric(.x))) %>%
    rename_with(
      ~ str_replace_all(.x, c(
        "w_w" = "ww",
        "^s" = "",
        "fettinnhold_percent" = "fat_pc",
        "lsi_percent" = "lsi_pc"
      )) %>%
        str_to_lower(),
      .cols = !matches("species")
    ) %>%
    # remove obs with no POP data
    filter(if_any(matches("ww"), ~ !is.na(.x))) %>%
    pivot_longer(cols = matches("ww"), names_to = "variable", values_to = "value") %>% 
  # remove outlier
  filter(lsi_pc < 15)
  
```


```{r}
cleaned_data %>% select(-variable, -value) %>% 
  mutate(across(!matches("species"), ~ as.numeric(.x))) %>% pivot_longer(cols = -"species", names_to = "variable", values_to = "value") %>% 
  
  
  ggplot(aes(value))+geom_histogram(bins = 100)+facet_grid(species~variable, scales = "free")
```




## Data description

* Observations:
  + Approx. 250 for each fish species (3) for each contaminant (6). Obs ikke like godt for alle.
* Response variables:
  + Concentrations of six different POPs in liver.
  + Both wet weight conc and lipid weight conc available.
* Explanatory variables:
  + Species (factor with three levels)
  + year (must be kept in model)
  + weight of fish (omitted)
  + weight of liver (omitted)
  + length of fish (log trans)
  + fat (percentage of liver)
  + lsi (liver weight vs total weight, nutritional condition)
  + age (missing variables, omitted)

From previous studies we know the effect of these variables will differ between species due to e.g.  different physiology and feeding habits. Hence, including fish as a factor in a single model will require an advanced model and is not justified unless a comparisons between the levels of the different fish species are important. Known confounders are fish age or proxies such as fish length, liver weight, hepatosomatic index and liver somatic index. LIVER FAT CONTENT?


## Model selection

* Main hypotheses:
  + **Is there a time trend for any of the contaminants in any of the species?**
* Descriptive analyses:
  + Which fish have the highest concentrations of each contaminant?
  + Is the effect of e.g. fat content different between species?

One identical model, fitted to each different dataset (3 species x  6 contaminants = 18). This is justified as included an additional explanatory variable should be inconsequential due to the large number of observations. Due to collinearity with lsi and fish length, weight of liver and fish was omitted. Fish age was omitted due to missingness of a substantial fraction of observations and collinearity with fish length. As such, a linear model including explanatory variables year, fish length, fat percentage and lsi was chosen.


## Model fitting

```{r model}
# linear regression model
pop_lm <- function(in_dat) {
  lm(
    data = in_dat, formula = value ~ year + length_cm + fat_pc + lsi_pc,
    na.action = na.exclude
  )
}

```


```{r model-fit-partial}

# conditional manual
p_man <- cleaned_data %>%
  ungroup() %>%
  group_by(species, variable) %>%
  drop_na(value) %>%
  nest() %>%
  # apply model and calculate medians of covariates
  mutate(
    model = map(data, pop_lm),
    med_length_cm = map(
      data,
      ~ median(.x$length_cm, na.rm = TRUE)
    ) %>% as.numeric(),
    med_fat_pc = map(
      data,
      ~ median(.x$fat_pc, na.rm = TRUE)
    ) %>% as.numeric(),
        med_lsi_pc = map(
      data,
      ~ median(.x$lsi_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_length_cm = map(
      data,
      ~ mean(.x$length_cm, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_fat_pc = map(
      data,
      ~ mean(.x$fat_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_lsi_pc = map(
      data,
      ~ mean(.x$lsi_pc, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_year = map(
      data,
      ~ mean(.x$year, na.rm = TRUE)
    ) %>% as.numeric(),
    mean_value = map(
      data,
      ~ mean(.x$value, na.rm = TRUE)
    ) %>% as.numeric()
  ) %>%
  mutate(
    glance = map(model, broom::glance),
    year_pval = map(model, ~ broom::tidy(.x)[[2, 5]] %>% as.numeric()),
    # tidy gets summary variables of the model coefficients, pivot to get coeffs for unnest
    tidy_coeff = map(model, ~ broom::tidy(.x) %>%
      select(term, estimate) %>%
      pivot_wider(names_from = "term", values_from = "estimate") %>%
      clean_names() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_coeff"))),
      tidy = map(model, ~ broom::tidy(.x)),
    # augment gets per observation variables (e.g., residuals)
    augment = map2(model, data, ~ broom::augment(.x, .y)),
    partial_resids = map(model, ~ residuals(.x, type = "partial") %>%
      as_tibble() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_partial_resid"))),
    normal_resids = map(model, ~ residuals(.x) %>%
      as_tibble() %>%
      rename_with(.cols = everything(), .fn = ~ paste0(.x, "_normal_resid")))
  ) %>%
  # add package-generated plots for an extra check
  mutate(
    visreg = map(
      model,
      ~ visreg(.x, "year", gg = TRUE, type = "conditional")
    ),
    effect_plot = map(
      model,
      ~ effect_plot(.x, pred = year, interval = TRUE, partial.residuals = TRUE)
    ),
    variable = toupper(variable) %>% str_replace_all("_WW", "")
  )

```

## Model validation

### Linearity

```{r}

#Residuals vs fitted and vs each covriate

p_man %>% unnest(augment) %>%
  ggplot(aes(.fitted, .resid)) +
    geom_point(size = 0.2) + 
    geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")

```


```{r}

#Residuals vs each covriate


p_man %>% unnest(augment) %>%
  ggplot(aes(length_cm, .resid)) +
    geom_point(size = 0.2) + 
      geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")

p_man %>% unnest(augment) %>%
  ggplot(aes(fat_pc, .resid)) +
    geom_point(size = 0.2) + 
      geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")

p_man %>% unnest(augment) %>%
  ggplot(aes(lsi_pc, .resid)) +
    geom_point(size = 0.2) + 
    geom_hline(yintercept = 0, color = "red") +
  geom_smooth(method = "loess", formula = y ~ x, se = T, span = 2)+
    facet_grid(species~variable, scales = "free")


```


Outliers in lsi found. Removed lsi > 20.

### Normality

#### Visual


##### QQ-plot of response

```{r}
ggplot(p_man %>% unnest(augment), aes(sample = value)) +
  stat_qq(size = 0.2) +
  stat_qq_line(color = "red", size = 0.4) +
  facet_grid(species ~ variable, scales = "free")
```

##### Distribution of response

```{r}
ggplot(p_man %>% unnest(augment), aes(value)) +
    geom_histogram(aes(y=..density..), color='gray50',
        alpha=0.2, position = "identity")+
    geom_density(alpha=0.2, fill = "red")+
  facet_grid(species ~ variable, scales = "free")
```

Bimodal, HCH indicating missing categorical.

##### Shapiro-Wilk's test


```{r shapiro-3col}
shap_table <- p_man %>% unnest(augment) %>% group_by(species, variable) %>% summarise(shapiro = shapiro.test(.std.resid) %>% tidy()) %>% unnest(shapiro) %>% select(-method, -statistic)

fish_table <- function(x){
  gt(x) %>% 
    tab_style(
    location = cells_body(
      columns = p.value,
      rows = p.value < 0.01
    ),
    style = list(cell_text(color = "red"))
  ) %>% fmt_number(columns = 2, decimals = 3) %>% 
    tab_options(column_labels.hidden = TRUE) %>% 
    as_raw_html() # return as html
}

left_table <- 
  shap_table %>% 
  filter(species == "Cod") %>% 
  ungroup() %>% 
  select(-species) %>% 
  fish_table()

mid_table <- 
  shap_table %>% 
  filter(species == "Haddock") %>%
    ungroup() %>% 
  select(-species) %>% 
  fish_table()

right_table <- 
  shap_table %>% 
  filter(species == "Saithe") %>%
    ungroup() %>% 
  select(-species) %>% 
  fish_table()

data_tables <- data.frame(left_table = left_table,
                          mid_table = mid_table,
                          right_table = right_table)

data_tables %>% 
  gt() %>% 
  fmt_markdown(columns = TRUE) %>% #render cell contents as html
  cols_label(left_table = "Cod",
             mid_table = "Haddock",
             right_table = "Saithe") %>% 
  tab_footnote("Red text indicates p-value < 0.01.")

```

### Homoskedasticity

#### Scale-location plot

```{r}
p_man %>%
  unnest(augment) %>%
  ggplot(aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point(size = 0.2) +
  geom_smooth(method = "loess", formula = y ~ x, span = 2, se = T, size = 0.5) +
  geom_smooth(method = "lm", formula = y ~ 1, se = FALSE, color = "red", size = 0.5, linetype="dashed", span = 2)+
  facet_grid(species ~ variable, scales = "free")
```

#### Breusch-Pagan test

```{r}
library(car)
 p_man %>%  mutate(
    b_pagan = map(
      model,
      ~ ncvTest(.x)$p
    )
) %>% select(species, variable, b_pagan) %>% ungroup() %>% gt()
```


### Independence
Location may be confounder, also time of year sampling took place. Different methodologies.

https://sscc.wisc.edu/sscc/pubs/RegDiag-R/index.html
https://vincentarelbundock.github.io/modelsummary/articles/modelsummary.html
https://thomaselove.github.io/431-notes/30-regr_diagnostics.html

Can run model w/o 2008 year.

## Partial residuals

### Background

Se: [@dunn_generalized_2018] p102 og [@sohil_introduction_2021][@moya-larano_plotting_2008]. "Partial residual plots could be used if the only purpose is to show how Y changes within the range of observed X values", and the slope of the fitted line corresponds to the regression coefficient in the complete model [@moya-larano_plotting_2008].

Alternativt og kanskje bedre: Gjøre Mann-Kendall test på residualer fra modell som korrigerer for lengde og fettinnhold. Plotte partielle residualer for GAM-modell. Da bruke nøyaktig samplingtid og gjøre MK-test plus avplot(?)
https://www.sciencedirect.com/science/article/pii/S1364815220310069
https://www.epa.gov/sites/default/files/2016-05/documents/tech_notes_6_dec2013_trend.pdf
Statistical Methods for Trend Detection and Analysis in the Environmental Sciences
https://cran.r-project.org/web/packages/trend/vignettes/trend.pdf

Assuming the least square estimate of the covariates $X_{1}$, $X_{2}$ and $X_{3}$ and $y_i$ and $\epsilon_i$ is the response and residual for the i-th point:

$$
y_i = f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}) + \epsilon_i.
$$

The partial residual of $X_3$ for the i-th observation is then generally defined as [@sohil_introduction_2021][@larsen_use_1972]:

$$
r_i = y_i-f_1\left(x_{i 1}\right)-f_2\left(x_{i 2}\right) = \epsilon_i + f_3(x_{i3}) ,
$$



$$
y_i = \hat b_0 + \hat b_{year}(year_{i})+ \hat b_{length}(length_{i})+\hat b_{fat pc}(fatperc_{i})+ \epsilon_i
$$
or, similarly

$$
 y_i = \bar y + \hat b_{year}\cdot (x_{year}-\bar x_{year}) + \hat b_{length}\cdot (x_{length}-\bar x_{length}) + \hat b_{fatpc}\cdot (x_{fatpc}-\bar x_{fatpc})  + \epsilon_i
$$


### Conditional partial residual plots

```{r}



p_man %>%
  unnest(tidy_coeff, partial_resids, data, normal_resids, augment) %>%
  ungroup() %>%
  ggplot(aes(year, (value_normal_resid + year * year_coeff + intercept_coeff + med_length_cm * length_cm_coeff + med_fat_pc * fat_pc_coeff + med_lsi_pc * lsi_pc_coeff))) +
  geom_jitter(size = 0.2, na.rm = T, height = 0, width = 0.2) +
  # geom_point(aes(year, value), size = 0.2, color = "darkgrey", position = position_jitternudge(width = 0.2, height = 0,
  #                                    seed = 123, x = 0.5,
  #                                    direction = "as.is",
  #                                    nudge.from = "jittered"))+
  ### TEST COND PART RESID
  # geom_point(aes(year, value_normal_resid + mean_value +
  #                year_coeff*(year - mean_year) + length_cm_coeff*(med_length_cm - mean_length_cm) + fat_pc_coeff*(med_fat_pc - mean_fat_pc)), size = 0.2, color = "orange", position = position_jitternudge(width = 0.2, height = 0,
  #                                    seed = 123, x = -0.5,
  #                                    direction = "as.is",
  #                                    nudge.from = "jittered"))+
  ### TEST COND PART RESID EN
  stat_smooth(method = "lm", na.rm = TRUE, se = TRUE, color = "red", size = 0.5) +
  # stat_smooth(method = "loess", na.rm = TRUE, se = TRUE, color = "orange", size = 0.4) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.key = element_blank(),
    strip.background = element_rect(colour="black", fill="white"),
    strip.text.x = element_text(face = "bold"),
    strip.text.y = element_text(face = "bold")
  ) +
  geom_text(
    data = p_man %>% group_by(species, variable) %>%
      distinct(year_pval, .keep_all = TRUE), color = "black",
    x = -Inf, y = Inf, hjust = -0.075, vjust = 1.5, size = 3,
    aes(label = paste0("p = ", round(year_pval %>% as.numeric(), digits = 3) %>% format(nsmall = 3)))
  ) +
  #   geom_text(
  #   data = p_man %>% unnest(tidy) %>% group_by(species, variable) %>%
  #     distinct(year_coeff, .keep_all = TRUE), color = "red",
  #   x = -Inf, y = Inf, hjust = -0.1, vjust = 3, size = 2.5,
  #   aes(label = paste0("slope = ", round(year_coeff %>% as.numeric(), digits = 4)))
  # ) +
  scale_x_continuous(breaks = c(2019, 2016, 2013, 2011, 2008, 2005, 2010)) +
  scale_y_continuous(breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4)) +
  ylab("log(Concentration + 1) [\u00b5g/kg]") +
  xlab("Year | (Length, Fat and LSI)") +

    facet_grid(species ~ factor(variable, levels = c(
    "PCB7",
    "DDT",
    "HCH",
    "HCB",
    "TNC",
    "PBDE"
  )))

ggsave("conditional_pr_stepan_data.png", width = 13, height = 7)

# comditional jtools and visreg
# p_man$visreg[[18]] + scale_y_continuous(breaks = pretty_breaks(n = 10))
# p_man$effect_plot[[18]]
# patchwork::wrap_plots(p_man$visreg)
```

## Strength of decrease

```{r}

p_man %>% clean_names() %>%
  select(species, variable, tidy) %>%
  unnest(tidy) %>%
  filter(term == "year") %>%
  select(species, variable, p.value, estimate) %>%
  mutate(estimate = 10^estimate - 1) %>%
  gt(groupname_col = "variable",
     rowname_col = "species") %>% 
  tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate > 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "red"))
  ) %>% 
  
    tab_style(
    location = cells_body(
      columns = estimate,
      rows = estimate < 0 & p.value < 0.05
    ),
    style = list(cell_text(color = "green"))
  ) %>% 
  
  
  fmt_number(everything(), decimals = 3) %>% 
  fmt_percent(estimate, decimals = 1) %>% 
  tab_footnote("Estimated annual percentage change obtained by exponentiation and corresponding p-values. Green indicates a significant decrease, whereas red indicates a significant increase.")





```



## Method description UPDATE
Statistical analyses were carried out in R version 4.2.1. Data with missing covariates or response variables were removed, and response variables were added one prior to logtransformation with base 10 to meet assumptions of homoscedasticity and normality while preserving zero-valued observations. The covariate length was also logtransformed. To correct for known confounders, a multiple linear regression model with the covariates year, length, fat percentage and liver somatic index was fitted  the cleaned and transformed data for each POP and species. To visually demonstrate the effect of year on the POP concentrations correcting for other dependent variables, the conditional partial residuals versus year was plotted, keeping length, fat and lsi constant at their median values. The significance of the regression coefficient for the year was determined using a Wald test to test the null hypothesis of no time trend.



## TODO

* Sjekke ut conditional partial residual plots etter modell er oppdatert.

* Oppdatere metodedescription.

* Breusch pagan fjerne. Evt. Levene i stedet?

* Fjerne shapiro wilk?

* Lage til supplementary

* DIskusjon at det er noen svakheter med statisik modell. I metodedel?



